---
title: "Homework 7: Factor Analysis"
subtitle: "S&DS 5360 | Multivariate Statistics"
author: "Brandon Tran (brandon.tran@yale.edu)"
date: today
date-format: "MMMM DD, YYYY"
format: pdf 
---



```{r, include = FALSE}
library(MASS)
library(dplyr)
library(ggplot2)
library(haven)
library(tidyr)
library(knitr)
library(kableExtra)
library(tidyverse)
library(corrplot)
library(EFAtools)
library(psych)
```



# Part 1 | Setup & Data Overview
We continue our work with the Healthy Minds Survey (HMS) data. For
Factor Analysis, we will select three distinct survey groups

- Depression (`phq9_1` to `phq9_9`): Nine questions on interest, mood, sleep, and energy levels over 2 weeks.
- Anxiety (`gad7_1` to `gad7_7`): Seven questions on nervousiness, worrying, and restlessness over 2 weeks.
- Flourishing (`diener1` to `diener8`): Eight questions assessing positive psychological and social well-being.
- Resilience (`brs_1` to `brs_6`): Six questions measuring a student's ability to cope with stress and adversity. Note that we will need to reverse-score the even questions so that the scale remains consistent (higher score indicates higher resilience).

We omit the code to preprocess the data.

```{r, echo = FALSE, cache = TRUE}
hms_data <- read_sav("../data/hms_data.sav", encoding = 'latin1')

fa_vars <- c(
  "phq9_1", "phq9_2", "phq9_3", "phq9_4", "phq9_5", "phq9_6", "phq9_7", "phq9_8", "phq9_9",
  "gad7_1", "gad7_2", "gad7_3", "gad7_4", "gad7_5", "gad7_6", "gad7_7",
  "diener1", "diener2", "diener3", "diener4", "diener5", "diener6", "diener7", "diener8",
  "BRS_1", "BRS_2", "BRS_3", "BRS_4", "BRS_5", "BRS_6"
)

df <- hms_data %>%
     select(all_of(fa_vars)) %>%
     drop_na() %>%
     mutate(
     BRS_2 = 6 - BRS_2,
     BRS_4 = 6 - BRS_4,
     BRS_6 = 6 - BRS_6
)
```



# Part 2 | Correlation Matrix
To better understand the relationship between our variables (the HMS survey
questions), we begin by presenting a correlation matrix. We expect corrleation,
particularly among within-group questions (e.g., different questions in the
`phq9` depression category should be correlated).

```{r}
cormat <- cor(df)
corrplot(cormat, method = 'color', type = 'upper',
         tl.col = 'black', tl.cex = 0.6,
         title = 'Correlation Matrix of HMS Indicators',
         mar = c(0,0,1,0))
```

As expected, observe our within-scale groupings (the dark blue triangular
blocks). The questions from `phq9`, `gad7`, and `diener` all exhibit
intra-group correlations, indicating strong internal consitency within the
scales. Turning our attention to the larger blue rectangles, these variables
represent the comorbidity between depression and anxiety indicators (`phq9` and
`gad7`, respectively). We have previously shown the interconnected relationship
between these two question groupings.

Finally, we see a large red rectangle emerge between the flourishing (`diener`) 
and resilience (`BRS`) questions with those of `phq9` and `gad7`. We have 
previously oberved this inverse relationship (or better said, negative 
correlation) in the past for the former: positive well-being and flourishing 
should be inversely related to psychological distress. It makese sense that
reslience would follow this same relationship.



# Part 3 | KMO for Suitability of Data for Analysis
To determine the suitablity of our data for factor analysis, we calculate the
Kaiser-Meyer-Olkin (KMO) measure of sampling adequacy using the `EFATools`
package introduced in class.

```{r}
KMO(as.matrix(df))
```

As seen above, the overall KMO score for our $30$ variables is "marvellously"
high at $0.959$. Further, note that the indivual scores for each variable
exceed $0.91$; thus, there is substantial shared variance among the questions,
making our data highly suitable for factor extraction.



# Step 4 | Determine Latent Factors
Next, to determine the appropriate number of latent factors to use for our 
analysis, we use both parallel analysis and our traditional scree plot and 
eigenvalue interpretation from principal components analysis.

```{r, fig.width = 8, fig.height = 4, width.out = '100%', cache = TRUE, warning = FALSE, message = FALSE}
# Parallel Analysis
PARALLEL(df)

# PCA with Scree Plot + Eigenvalues
comp1 <- prcomp(df, scale. = TRUE)

screeplot(comp1, type = 'lines', col = 'red', lwd = 2, pch = 19, cex = 1.2,
          main = 'Scree Plot of HMS Indicators')

eigenvalues <- comp1$sdev^2
round(eigenvalues[1:10], 2)
```

Observe that the PCA-based decision rule suggests 4 factors (this is where the
real eigenvalues drop below the simulated 95th percentile threshold). Meanwhile,
SMC and EFA suggest 11 and 7, respectively. We turn to the scree plot, which has
a distinct elbow at 4 factors. Considering after $4$, our eigenvalues drop
below $1.0$, as well as the fact that our data come from four distinct
measurement groups (`PHQ`, `GAD`, etc.), it makes the most sense to proceed with
$4$ latent factors.



# Part 5 | Factor Analyses
We now perform a series of factor analyses using orthogonal models with 4 latent
factors. We elect to compare all four presented extraction methods: Principal
Components, Principal Axis Factoring, Iterative PCA, and Maximum Likelihood. We
will evaluate each method based on their root mean squared residual (RSMR),
focusing on the percentage of residuals that exceed $0.05$.

```{r}
# Principal Components Analysis (PCA)
fact_pca <- principal(df, nfactors = 4, rotate = "varimax")
repro_pca <- fact_pca$loadings %*% t(fact_pca$loadings)
resid_pca <- cor(df) - repro_pca
len_pca <- length(resid_pca[upper.tri(resid_pca)])
rmsr_pca <- sqrt(sum(resid_pca[upper.tri(resid_pca)]^2) / len_pca)
prop_pca <- sum(abs(resid_pca[upper.tri(resid_pca)]) > 0.05) / len_pca * 100

# Principal Axis Factoring (PAF)
fact_paf <- fa(df, nfactors = 4, rotate = "varimax", fm = "pa")
repro_paf <- fact_paf$loadings %*% t(fact_paf$loadings)
resid_paf <- cor(df) - repro_paf
len_paf <- length(resid_paf[upper.tri(resid_paf)])
rmsr_paf <- sqrt(sum(resid_paf[upper.tri(resid_paf)]^2) / len_paf)
prop_paf <- sum(abs(resid_paf[upper.tri(resid_paf)]) > 0.05) / len_paf * 100

# Iterative PCA
fact_ipca <- fa(df, nfactors = 4, rotate = "varimax", SMC = FALSE, fm = "pa")
repro_ipca <- fact_ipca$loadings %*% t(fact_ipca$loadings)
resid_ipca <- cor(df) - repro_ipca
len_ipca <- length(resid_ipca[upper.tri(resid_ipca)])
rmsr_ipca <- sqrt(sum(resid_ipca[upper.tri(resid_ipca)]^2) / len_ipca)
prop_ipca <- sum(abs(resid_ipca[upper.tri(resid_ipca)]) > 0.05) / len_ipca * 100

# Maximum Likelihood
fact_ml <- factanal(df, factors = 4, rotation = "varimax")
repro_ml <- fact_ml$loadings %*% t(fact_ml$loadings)
resid_ml <- cor(df) - repro_ml
len_ml <- length(resid_ml[upper.tri(resid_ml)])
rmsr_ml <- sqrt(sum(resid_ml[upper.tri(resid_ml)]^2) / len_ml)
prop_ml <- sum(abs(resid_ml[upper.tri(resid_ml)]) > 0.05) / len_ml * 100
```

```{r, echo = FALSE}
extraction_results <- data.frame(
  Method = c("Maximum Likelihood (ML)", 
             "Principal Axis Factoring (PAF)", 
             "Iterative PCA", 
             "Principal Components (PCA)"),
  RMSR = c(0.0269, 0.0260, 0.0260, 0.0392),
  Residuals = c("7.59%", "6.67%", "6.67%", "14.25%")
)

# 2. Rename columns for the table output
colnames(extraction_results) <- c("Extraction Method", "RMSR", "% Residuals > 0.05")

# 3. Generate the kable table formatted for LaTeX/PDF
kable(extraction_results, format = "latex", booktabs = TRUE, 
      caption = "Comparison of Factor Extraction Methods (4-Factor Model)",
      align = c("l", "c", "c")) %>%
  kable_styling(latex_options = c("HOLD_position", "striped"), 
                full_width = FALSE, 
                position = "center") %>%
  row_spec(0, bold = TRUE) %>%
  # Highlight rows 2 and 3 (PAF and Iterative PCA) since they tied for the best fit
  row_spec(c(2, 3), bold = TRUE)
```

We see that PCA performs the worst, yielding the highest RMSR of $0.0392$ and
a substantial $14.25\%$ of residuals above the $0.05$ threshold. This matches
our expecations, given PCA analyzes total variance instead of isolating shared
latent variance. Meanwhile, maximum likelihood yielded improved performance,
with an RMSR of $0.0269$ and only $7.59\%$ above the threshold.

Both principal axis factoring and iterative PCA converged to the same limits,
and thus, they both are our strongest performers, with an RMSR of $0.0260$ and
$6.67\% > 0.05$. We elect to proceed with principal axis factoring given its
more commonplace usage with psychological constructs like ours, but we would
feel equally comfortable continuing with iterative PCA.



# Part 6 | Rotations
Now, we explore two rotational methods: varimax, which maximizes the variance
of squared loadings (encouraging each latent factor to be comprised of distinctly
correlated questions), and quartimax, which minimizes the number of factors
needed to explain each variable (encouraging incorporation of as many questions
as possible on to the same latent factor).

```{r, warning = FALSE, message = FALSE}
# Varimax
paf_varimax <- fa(df, nfactors = 4, rotate = 'varimax', fm = 'pa')
# Quartimax
paf_quartimax <- fa(df, nfactors = 4, rotate = 'quartimax', fm = 'pa')
```

```{r, echo = FALSE}
loadings_varimax <- unclass(paf_varimax$loadings)
loadings_quartimax <- unclass(paf_quartimax$loadings)

# 2. Bind them side-by-side into one large matrix
combined_loadings <- cbind(loadings_varimax, loadings_quartimax)

# 3. Clean the data: Round to 3 decimals and hide weak correlations (< 0.3)
combined_loadings <- round(combined_loadings, 3)
combined_loadings[abs(combined_loadings) < 0.3] <- ""

# 4. Convert to a data frame so kable can read it
df_loadings <- as.data.frame(combined_loadings)

# Rename the columns so they are clean (4 for Varimax, 4 for Quartimax)
colnames(df_loadings) <- c("PA1", "PA2", "PA3", "PA4", "PA1", "PA2", "PA3", "PA4")

# 5. Generate the LaTeX-formatted PDF table
kable(df_loadings, format = "latex", booktabs = TRUE, 
      caption = "Comparison of Varimax and Quartimax Factor Loadings",
      align = "c") %>%
  kable_styling(latex_options = c("scale_down", "striped", "HOLD_position"), 
                font_size = 9) %>%
  # This adds the overarching super-headers!
  add_header_above(c(" " = 1, "Varimax Rotation" = 4, "Quartimax Rotation" = 4)) %>%
  # Make the variable names bold
  column_spec(1, bold = TRUE) %>%
  # Add a vertical divider line between the Varimax and Quartimax sections
  column_spec(5, border_right = TRUE)
```

Observe under quartimax that for `PA1`, almost every single `phq9` and `gad7`
question was loaded, combining depression and anxiety into the same latent
factor. As a result, we see that `PA4` has become almost useless, explaining
only $3\%$ of the variance. Meanwhile, for the varimax loadings, we see
largely separable columns:

- `PA1` contains the flourishing (`diener`) questions
- `PA2` contains the `phq9` (depression) questions
- `PA3` contains the `gad7` (anxiety) questions, and
- `PA4` contains the `BRS` (resilience) questions.

As expected, there is overlap between `PA2` and `PA3` given the comorbidity of
depression and anxiety.

```{r, fig.width = 8, fig.height = 4, width.out = '100%', cache = TRUE}
par(mfrow = c(1,2))

# Varimax
plot(paf_varimax$loadings[, c(1, 2)], pch = 18, col = 'red', 
     main = "Varimax (PA1 vs PA2)", xlim = c(-1, 1), ylim = c(-1, 1))
abline(h = 0, lty = 2); abline(v = 0, lty = 2)
text(paf_varimax$loadings[, c(1, 2)], labels = rownames(paf_varimax$loadings), 
     cex = 0.7, pos = 4, col = "blue")

# Quartimax
plot(paf_quartimax$loadings[, c(1, 2)], pch = 18, col = 'red', 
     main = "Quartimax (PA1 vs PA2)", xlim = c(-1, 1), ylim = c(-1, 1))
abline(h = 0, lty = 2); abline(v = 0, lty = 2)
text(paf_quartimax$loadings[, c(1, 2)], labels = rownames(paf_quartimax$loadings), 
     cex = 0.7, pos = 4, col = "darkgreen")
```

The loading plots confirm this distinction. In the varimax plot of PA1 versus
PA2, the variable clusters tightly hug the axes (e.g., flourishing questions 
load along the vertical axis). In contrast, the quartimax plot shows variables
floating somewhat ambiguously within the quadrants, visually demonstrating
cross-loadings across multiple dimensions. Thus, we proceed with the varimax
method, presenting all four dimensions via a pairs plot:

```{r, fig.width = 8, fig.height = 8, width.out = '100%', cache = TRUE}
pairs(paf_varimax$loadings, 
      main = "Varimax Loadings Matrix (All 4 Dimensions)",
      pch = 18, col = "blue", cex = 1.2)
```

We see groupings of questions sitting along the vetical and horizontal axes,
indicating clear separation and simple structure. Even with both depression
(PA2) and anxiety (PA3), which we know are correlated variables, we see that
these have been separated into distinct dimensions, give or take some slight
dispersion that reflects the known comorbidity between the two conditions.



# Part 7 | Summary
Our exploratory factor analysis successfully identified the latent constructs
driving mental health and well-being indicators among the students represented
in the 2024–25 Healthy Minds Survey. Initial diagnostics indicated excellent
sampling adequacy, with an overall KMO of $0.959$. Subsequently, parallel
analysis indicated the presence of four distinct latent factors. A comparison
of four orthogonal models via RMSR and two rotational methods brought us to
decide to utilize principal axis factoring with a varimax rotation.

This approach cleanly resolved our 30 questions into four distinct dimensions that
directly map to the questions coded categories: depression, flourishing, anxiety,
and resilience. While slight cross-loadings between anxiety and depression reflect
their known clinical comorbidity, the model demonstrates that students' mental
health profiles are well structured along these four distinct axes of distress
and well-being.