---
title: "Homework 7: Factor Analysis"
subtitle: "S&DS 5360 | Multivariate Statistics"
author: "Brandon Tran (brandon.tran@yale.edu)"
date: today
date-format: "MMMM DD, YYYY"
format: pdf 
---



```{r, include = FALSE}
library(MASS)
library(dplyr)
library(ggplot2)
library(haven)
library(tidyr)
library(knitr)
library(kableExtra)
library(tidyverse)
library(corrplot)
library(EFAtools)
```



# Part 1 | Setup & Data Overview
We continue our work with the Healthy Minds Survey (HMS) data. For
Factor Analysis, we will select three distinct survey groups

- Depression (`phq9_1` to `phq9_9`): Nine questions on interest, mood, sleep, and energy levels over 2 weeks.
- Anxiety (`gad7_1` to `gad7_7`): Seven questions on nervousiness, worrying, and restlessness over 2 weeks.
- Flourishing (`diener1` to `diener8`): Eight questions assessing positive psychological and social well-being.
- Resilience (`brs_1` to `brs_6`): Six questions measuring a student's ability to cope with stress and adversity. Note that we will need to reverse-score the even questions so that the scale remains consistent (higher score indicates higher resilience).

We omit the code to preprocess the data.

```{r, echo = FALSE, cache = TRUE}
hms_data <- read_sav("data/hms_data.sav", encoding = 'latin1')

fa_vars <- c(
  "phq9_1", "phq9_2", "phq9_3", "phq9_4", "phq9_5", "phq9_6", "phq9_7", "phq9_8", "phq9_9",
  "gad7_1", "gad7_2", "gad7_3", "gad7_4", "gad7_5", "gad7_6", "gad7_7",
  "diener1", "diener2", "diener3", "diener4", "diener5", "diener6", "diener7", "diener8",
  "BRS_1", "BRS_2", "BRS_3", "BRS_4", "BRS_5", "BRS_6"
)

df <- hms_data %>%
     select(all_of(fa_vars)) %>%
     drop_na() %>%
     mutate(
     BRS_2 = 6 - BRS_2,
     BRS_4 = 6 - BRS_4,
     BRS_6 = 6 - BRS_6
)
```



# Part 2 | Correlation Matrix
To better understand the relationship between our variables (the HMS survey
questions), we begin by presenting a correlation matrix. We expect corrleation,
particularly among within-group questions (e.g., different questions in the
`phq9` depression category should be correlated).

```{r}
cormat <- cor(df)
corrplot(cormat, method = 'color', type = 'upper',
         tl.col = 'black', tl.cex = 0.6,
         title = 'Correlation Matrix of HMS Indicators',
         mar = c(0,0,1,0))
```

As expected, observe our within-scale groupings (the dark blue triangular
blocks). The questions from `phq9`, `gad7`, and `diener` all exhibit
intra-group correlations, indicating strong internal consitency within the
scales. Turning our attention to the larger blue rectangles, these variables
represent the comorbidity between depression and anxiety indicators (`phq9` and
`gad7`, respectively). We have previously shown the interconnected relationship
between these two question groupings.

Finally, we see a large red rectangle emerge between the flourishing (`diener`) 
and resilience (`BRS`) questions with those of `phq9` and `gad7`. We have 
previously oberved this inverse relationship (or better said, negative 
correlation) in the past for the former: positive well-being and flourishing 
should be inversely related to psychological distress. It makese sense that
reslience would follow this same relationship.



# Part 3 | KMO for Suitability of Data for Analysis
To determine the suitablity of our data for factor analysis, we calculate the
Kaiser-Meyer-Olkin (KMO) measure of sampling adequacy using the `EFATools`
package introduced in class.

```{r}
KMO(as.matrix(df))
```

As seen above, the overall KMO score for our $30$ variables is "marvellously"
high at $0.959$. Further, note that the indivual scores for each variable
exceed $0.91$; thus, there is substantial shared variance among the questions,
making our data highly suitable for factor extraction.



# Step 4 | Determine Latent Factors
Next, to determine the appropriate number of latent factors to use for our 
analysis, we use both parallel analysis and our traditional scree plot and 
eigenvalue interpretation from principal components analysis.

```{r, fig.width = 8, fig.height = 6, width.out = '100%', cache = TRUE}
# Parallel Analysis
PARALLEL(df)

# PCA with Scree Plot + Eigenvalues
comp1 <- prcomp(df, scale. = TRUE)

screeplot(comp1, type = 'lines', col = 'red', lwd = 2, pch = 19, cex = 1.2,
          main = 'Scree Plot of HMS Indicators')

eigenvalues <- comp1$sdev^2
round(eigenvalues[1:10], 2)
```

Observe that the PCA-based decision rule suggests 4 factors (this is where the
real eigenvalues drop below the simulated 95th percentile threshold). Meanwhile,
SMC and EFA suggest 11 and 7, respectively. We turn to the scree plot, which has
a distinct elbow at 4 factors. Considering after $4$, our eigenvalues drop
below $1.0$, as well as the fact that our data come from four distinct
measurement groups (`PHQ`, `GAD`, etc.), it makes the most sense to proceed with
$4$ latent factors.

