---
title: "Final Project: Code for Report"
subtitle: "S&DS 5630 | Multivariate Statistics"
author: "Brandon Tran (brandon.tran@yale.edu)"
date: today
date-format: "MMMM DD, YYYY"
format: pdf 
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load required libraries
library(tidyverse)
library(dplyr)
library(haven)
library(corrplot)
library(MASS)
library(car)
library(NbClust)
library(heplots)
library(vegan)
library(PerformanceAnalytics)
```



# Variables
```{r, cache=TRUE}
# Load Data
hms_data <- read_sav("data/hms_data.sav", encoding = 'latin1')

symptoms <- c(paste0('phq9_', 1:9), paste0('gad7_', 1:7))
environment <- c('age', 'lonesc', 'exerc', 'internet_1', 'fincur')

# Data Cleaning
df <- hms_data %>%
  mutate(across(c(gender_trans, gender_queer, gender_nonbin, gender_selfID,
                  sexual_h, sexual_l, sexual_g, sexual_bi, sexual_queer,
                  sexual_quest, sexual_pan, sexual_asexual),
                  ~replace_na(., 0))) %>%
  mutate(
      trans = (gender_trans == 1 | gender_queer == 1 | gender_nonbin == 1 | 
               gender_selfID == 1),
      queer_flag = (sexual_l == 1 | sexual_g == 1 | sexual_bi == 1 | 
                    sexual_queer == 1 | sexual_quest == 1 | sexual_pan == 1 | 
                    sexual_asexual == 1),
      identity_group = case_when(
          trans ~ "TGNC",
          !trans & queer_flag ~ "Cis-LGBQ+",
          !trans & !queer_flag & sexual_h == 1 ~ "Cis-Het",
          TRUE ~ NA_character_
      )
  ) %>%
  mutate(
      identity_group = as.factor(identity_group),
      ther_any = as.factor(ther_any),
      queer = as.factor(queer_flag),
      sleep_wknight = as.numeric(sleep_wknight),
      deprawsc = as.numeric(deprawsc),
      flourish = as.numeric(flourish)
  )
```



# Descriptive Plots and Transformations
```{r}
df %>%
  dplyr::select(flourish, deprawsc, anx_score, lonesc, ed_sde) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Score") %>%
  ggplot(aes(x = Score)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  facet_wrap(~ Variable, scales = "free") +
  theme_minimal() +
  labs(title = "Histograms of Mental Health Variables",
       x = "Score", y = "Frequency")

ggsave("Final Project/figures/4.1.1_histograms.pdf", width = 7, height = 5)
```

```{r}
pca_data <- df %>% 
  dplyr::select(flourish, deprawsc, anx_score, lonesc, ed_sde) %>%
  drop_na()

corr_matrix <- cor(pca_data)
pdf("Final Project/figures/4.1.2_corrplot.pdf", width = 8, height = 8)
corrplot.mixed(corr_matrix, lower = "number", upper = "ellipse")
dev.off()
```

```{r}
set.seed(4747)
pca_plot_sample <- pca_data %>% slice_sample(n = 5000)

pdf("Final Project/figures/4.1.3_matrixplot.pdf", width = 8, height = 8)
chart.Correlation(pca_plot_sample, histogram = TRUE, pch = 19)
dev.off()
```

```{r}
pdf("Final Project/figures/4.4_q-q.pdf", width = 7, height = 5)
cqplot(pca_data, main = "Chi-Square Q-Q Plot for Raw Data")
dev.off()
```

```{r}
df_transform <- df %>%
    mutate(
        deprawsc_sqrt = sqrt(deprawsc + 0.5),
        anx_sqrt = sqrt(anx_score + 0.5),
        flourish_sq = flourish^2
    )

pca_data_transform <- df_transform %>% 
  dplyr::select(flourish_sq, deprawsc_sqrt, anx_sqrt, lonesc, ed_sde) %>%
  drop_na()

pdf("Final Project/figures/4.5_post-transform.pdf", width = 7, height = 5)
cqplot(pca_data_transform, main = "Chi-Square Q-Q Plot for Transformed Data")
dev.off()
```

```{r, echo=FALSE}
# Original correlation matrix
corr_matrix_orig <- cor(pca_data, use = "pairwise.complete.obs")

# Transformed correlation matrix
corr_matrix_trans <- cor(pca_data_transform, use = "pairwise.complete.obs")

# Extract lower triangles for comparison
lower_tri_idx <- lower.tri(corr_matrix_orig)

comparison_table <- data.frame(
  Pair = paste(rownames(corr_matrix_trans)[row(corr_matrix_trans)[lower_tri_idx]], 
               "vs", 
               colnames(corr_matrix_orig)[col(corr_matrix_orig)[lower_tri_idx]]),
  Original = corr_matrix_orig[lower_tri_idx],
  Transformed = corr_matrix_trans[lower_tri_idx]
)

comparison_table$Change <- comparison_table$Transformed - comparison_table$Original
```



# PCA Dimensionality Check
```{r pca-analysis, message=FALSE, warning=FALSE}
# 1. Run PCA on the scaled, transformed data
pca_model <- prcomp(pca_data_transform, center = TRUE, scale. = TRUE)

# 2. Extract Variance & Eigenvalues for Table
pca_summary <- summary(pca_model)
eigenvalues <- pca_model$sdev^2
variance_explained <- pca_summary$importance[2, ] * 100
cumulative_variance <- pca_summary$importance[3, ] * 100

cat("--- PCA Variance Summary ---\n")
pca_table <- data.frame(
  Eigenvalue = round(eigenvalues, 3),
  Variance_Pct = round(variance_explained, 1),
  Cumulative_Pct = round(cumulative_variance, 1)
)
print(pca_table)

cat("\n--- PCA Loadings (PC1 & PC2) ---\n")
print(round(pca_model$rotation[, 1:2], 3))

# 3. Generate PCA Biplot
library(factoextra)
pdf("Final Project/figures/4.6_pca_biplot.pdf", width = 7, height = 5)
pca_plot <- fviz_pca_var(pca_model,
             col.var = "contrib", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE,     
             title = "PCA Biplot: Variable Contributions")
print(pca_plot)
dev.off()
```



# Ordination
```{r, message=FALSE, warning=FALSE, results='hide'}
# 1. Subset Data for Ordination
# (Sampling is required for computational feasibility and to avoid overplotting)
set.seed(4747)
df_subset <- df %>% 
  dplyr::select(all_of(symptoms), all_of(environment)) %>%
  drop_na() %>%
  sample_n(500)

hms_symptoms_subset <- df_subset %>% dplyr::select(all_of(symptoms))
hms_env_subset <- df_subset %>% dplyr::select(all_of(environment))

# 2. Correspondence Analysis (CA)
hms_ca <- cca(hms_symptoms_subset)

pdf("Final Project/figures/5.1_ca_plot.pdf", width = 7, height = 5)
plot(hms_ca, choices = c(1, 2), type = "n", 
     main = "CA of Symptoms (Dimensions 1 & 2)")
points(hms_ca, display = "sites", col = "lightgrey", pch = 16, cex = 0.5)
text(hms_ca, display = "species", col = "blue", cex = 1, font = 2)
dev.off()

# 3. Non-Metric Multidimensional Scaling (NMDS)
hms_nmds1 <- metaMDS(hms_symptoms_subset, distance = "bray", k = 1, trymax = 50)
hms_nmds2 <- metaMDS(hms_symptoms_subset, distance = "bray", k = 2, trymax = 50)
hms_nmds3 <- metaMDS(hms_symptoms_subset, distance = "bray", k = 3, trymax = 50)

pdf("Final Project/figures/5.1_nmds_screeplot.pdf", width = 7, height = 5)
stress_1d <- hms_nmds1$stress
stress_2d <- hms_nmds2$stress
stress_3d <- hms_nmds3$stress

dimensions <- c(1, 2, 3)
stress_values <- c(stress_1d, stress_2d, stress_3d)

plot(dimensions, stress_values, type = "b", pch = 19, col = "blue", lwd = 2,
     xlab = "Number of Dimensions (k)", ylab = "Stress",
     main = "NMDS Scree Plot for HMS Symptoms",
     ylim = c(0, max(stress_values) + 0.05), xaxt = "n")
axis(1, at = c(1, 2, 3))
abline(h = 0.20, col = "red", lty = 2)
text(2.5, 0.21, "Acceptable Threshold (0.20)", col = "red", cex = 0.8)
dev.off()

# 4. Environmental Vectors Overlay (Envfit)
fit_env <- envfit(hms_nmds2, hms_env_subset, permutations = 999)

pdf("Final Project/figures/5.1_nmds_envfit.pdf", width = 7, height = 5)
plot(hms_nmds2, type = "n", main = "NMDS with Environmental Vectors")
points(hms_nmds2, display = "sites", col = "lightgrey", pch = 16, cex = 0.5)
text(hms_nmds2, display = "species", col = "blue", cex = 0.9, font = 2)
plot(fit_env, col = "red", lwd = 2)
dev.off()

# 5. Non-Linear GAM Surface Wireplot (Ordisurf)
pdf("Final Project/figures/5.1_nmds_ordisurf.pdf", width = 7, height = 5)
plot(hms_nmds2, type = "n", main = "NMDS Wireplot: Loneliness Gradient")
points(hms_nmds2, display = "sites", col = "lightgrey", pch = 16, cex = 0.5)
text(hms_nmds2, display = "species", col = "blue", cex = 0.9, font = 2)
surf_lone <- ordisurf(hms_nmds2, hms_env_subset$lonesc, add = TRUE,
                      col = "darkgreen", lwd = 2)
dev.off()
```



# Two-Way Manova
```{r manova-analysis, message=FALSE, warning=FALSE}
# 1. Subset Data for MANOVA
df_manova <- df %>% 
  dplyr::select(deprawsc, flourish, ther_any, queer, sleep_wknight) %>%
  drop_na()

# 2. Main Effects & Interaction (Type III MANOVA)
manova_fit <- lm(cbind(deprawsc, flourish) ~ ther_any * queer, data = df_manova)
cat("--- Type III MANOVA ---\n")
print(Manova(manova_fit, type = 'III'))

# Univariate ANOVA (Generates F-Stats for Table 3 / tab:manova_main_effects)
cat("\n--- Univariate ANOVAs ---\n")
summary.aov(manova_fit)

# 3. Estimated Marginal Means (Contrasts)
library(emmeans)
emm_queer <- emmeans(manova_fit, ~ queer | rep.meas)
emm_ther <- emmeans(manova_fit, ~ ther_any | rep.meas)

cat("\n--- Queer Contrasts ---\n")
summary(pairs(emm_queer), adjust = 'none')

cat("\n--- Therapy Contrasts ---\n")
summary(pairs(emm_ther), adjust = 'none')

# 4. Covariate Integration: Sleep
model_cov <- lm(cbind(deprawsc, flourish) ~ ther_any + queer + sleep_wknight, data = df_manova)
cat("\n--- MANOVA with Sleep Covariate ---\n")
print(Manova(model_cov, type = "III"))

# 5. Evaluate Assumptions (Chi-Square Q-Q Plot)
# Note: Using model_cov as evaluated in the write-up
pdf("Final Project/figures/5.2_manova_qq.pdf", width = 7, height = 5)
cqplot(model_cov, main = "Chi-Square Q-Q Plot: Multivariate Normality Check")
dev.off()

# 6. Non-Parametric Validation (MRPP)
# Sampling is required for computational feasibility
set.seed(4747)
df_subset_mrpp <- df_manova[sample(nrow(df_manova), 2000), ]
dist_matrix <- vegdist(cbind(df_subset_mrpp$deprawsc, df_subset_mrpp$flourish), method = 'euclidean')

cat("\n--- MRPP: Sexual Orientation ---\n")
mrpp(dist_matrix, grouping = df_subset_mrpp$queer)

cat("\n--- MRPP: Therapy Utilization ---\n")
mrpp(dist_matrix, grouping = df_subset_mrpp$ther_any)
```

# Method 3: Discriminant Analysis
```{r da-analysis, message=FALSE, warning=FALSE}
# 1. Subset Data for DA
df_da <- df %>%
  dplyr::select(identity_group, flourish, deprawsc, anx_score, lonesc, ed_sde) %>%
  drop_na() %>%
  filter(identity_group %in% c("Cis-Het", "Cis-LGBQ+", "TGNC")) %>%
  mutate(identity_group = droplevels(identity_group))

# 2. Box's M Test
library(heplots)
cat("--- Box's M Test ---\n")
boxM(df_da[, -1], df_da$identity_group)

# 3. Model Comparison (CV = TRUE)
library(MASS)
lda_cv <- lda(identity_group ~ ., data = df_da, CV = TRUE)
qda_cv <- qda(identity_group ~ ., data = df_da, CV = TRUE)

acc_lda <- mean(lda_cv$class == df_da$identity_group)
acc_qda <- mean(qda_cv$class == df_da$identity_group)

cat("\n--- Classification Accuracy ---\n")
cat("LDA Accuracy:", round(acc_lda, 4), "\n")
cat("QDA Accuracy:", round(acc_qda, 4), "\n")

# 4. Stepwise Selection
library(klaR)
cat("\n--- Stepwise Discriminant Analysis ---\n")
# stepclass uses cross-validation to select variables.
# Note: as.data.frame() is required here because klaR does not support tibbles!
step_da <- stepclass(identity_group ~ ., data = as.data.frame(df_da), method = "lda", direction = "both")
print(step_da)

# 5. Final LDA Model (for coefficients and trace)
lda_model <- lda(identity_group ~ ., data = df_da)
cat("\n--- Final LDA Model Summary ---\n")
print(lda_model)

# Calculate Standardized Coefficients (Scaling * diag(sqrt(diag(cov(data)))))
sdev <- apply(df_da[, -1], 2, sd)
std_coef <- diag(sdev) %*% lda_model$scaling
rownames(std_coef) <- colnames(df_da)[-1]
cat("\n--- Standardized Coefficients ---\n")
print(std_coef)

# 6. Confusion Matrix & Recall
cat("\n--- Confusion Matrix (LDA CV) ---\n")
conf_matrix <- table(True = df_da$identity_group, Predicted = lda_cv$class)
print(conf_matrix)

cat("\n--- Class Recall (Sensitivity) ---\n")
recall <- diag(conf_matrix) / rowSums(conf_matrix)
print(round(recall, 4))

# 7. Visualizations: Score Plot
# Get predictions to plot LD1 and LD2
lda_predict <- predict(lda_model)
df_lda_plot <- data.frame(df_da, lda_predict$x)

library(ggplot2)
pdf("Final Project/figures/5.3_lda_scoreplot.pdf", width = 7, height = 5)
score_plot <- ggplot(df_lda_plot, aes(x = LD1, y = LD2, color = identity_group)) +
  geom_point(alpha = 0.3, size = 0.8) +
  stat_ellipse(linewidth = 1.2) +
  scale_color_manual(values = c("Cis-Het" = "#1b9e77", "Cis-LGBQ+" = "#d95f02", "TGNC" = "#7570b3")) +
  theme_minimal() +
  labs(title = "LDA Score Plot: Identity Groups", 
       x = "Linear Discriminant 1", y = "Linear Discriminant 2", color = "Identity Group")
print(score_plot)
dev.off()

# 8. Visualizations: Partimat Plot
# Note: Partimat is very computationally heavy. We take a stratified sample to make compilation fast and the plot readable.
set.seed(4747)
df_da_sample <- df_da %>% 
  group_by(identity_group) %>% 
  sample_n(150) %>% 
  ungroup()

pdf("Final Project/figures/5.3_lda_partimat.pdf", width = 8, height = 8)
partimat(identity_group ~ flourish + deprawsc + anx_score + lonesc, 
         data = df_da_sample, method = "lda", 
         image.colors = c("lightgreen", "lightcoral", "lightblue"))
dev.off()