---
title: "Final Project: Consolidated Draft"
subtitle: "S&DS 5630 | Multivariate Statistics"
author: "Brandon Tran (brandon.tran@yale.edu)"
date: today
date-format: "MMMM DD, YYYY"
format: pdf 
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load required libraries
library(tidyverse)
library(dplyr)
library(haven)
library(corrplot)
library(MASS)
library(car)
library(NbClust)
library(heplots)
library(vegan)
```


# Introduction
Over the course of the semester, we studied the Healthy Minds Network's latest 2024–2025 Healthy Minds Study (HMS) student survey. The HMS is a population-level, web-based survey administered by the HMN to assess mental health, service utilization, and related factors among post-secondary students. Since its inception over 15 years ago, the study has collected more than $935,000$ responses from over $675$ institution nationwide. The proposed research aims to examine the multivariate relationship between sexual and gender minority (GSM) identities, digital environment exposure, and mental health outcomes in the 2024-25 college student population. Utilizing the Demographics and Mental Health Status standard modules, this study will investigate how latent factors of psychological distress—comprising the PHQ-9 (Depression), GAD-7 (Anxiety), and UCLA Loneliness Scale—vary across intersectional identities.



# Planned Analysis (Edit)
To understand the latent structures and relationships within this data, we will perform four multivariate techniques:

Principal Component Analysis (PCA): To reduce our correlated mental health variables into a single index of psychological distress.

Ordination (CA & NMDS): To non-linearly map the topological structure of specific, individual mental health symptoms and evaluate how they correlate with environmental stressors.

MANOVA: To evaluate how specific demographics (sexual orientation and therapy utilization) impact these mental health outcomes simultaneously.

Discriminant Analysis: To determine if these psychological profiles can accurately predict a student's gender and sexual identity group.



# Variables
The resulting dataset features $n = 84,735$ individual respondents. For the purpose of this analysis, we will focus on questions that fall under the Standard Modules. Specifically, we will use composite continuous variables, individual symptom frequencies, and categorical predictors:

- Composite Scores: flourish (8-56), deprawsc (0-27), anx_score (0-21), lonesc (3-9), and ed_sde (0-5).

    - flourish: (Continuous) 8 statements using a 1-7 scale to indicate agreement with items like "I lead a purposeful and meaningful life." (min = 8, max = 56) .

    - dprawsc: (Continuous) 9 questions over the last 2 weeks regarding problems like "Feeling down, depressed or hopeless." (min = 0, max = 27) .

    - anx_score: (Continuous) 7 questions over the last 2 weeks regarding problems like "Feeling nervous, anxious or on edge." (min = 0, max = 21) .

    - lonesc: (Continuous) 3 questions regarding lacking companionship and feeling isolated. (min = 3, max = 9) .

    - ed_sde: (Continuous) 5 questions regarding disordered eating desires and behaviors. (min = 0, max = 5) .

- Individual Symptoms (Ordinal 1-4): Depression (phq9_1 to phq9_9) and Anxiety (gad7_1 to gad7_7).

- Environmental/Lifestyle: exerc (exercise hours), internet_1 (time online), fincur (financial stress), and sleep_wknight (hours of sleep).

- Identity Groups / Categorical Predictors: queer (Binary 0/1 for sexual orientation), ther_any (Binary 0/1 for therapy utilization), and identity_group (Categorical: Cis-Het, Cis-LGBQ+, TGNC).

```{r, cache=TRUE}
# Load data
hms_data <- read_sav("../data/hms_data.sav", encoding = 'latin1')

symptoms <- c(paste0('phq9_', 1:9), paste0('gad7_', 1:7))
environment <- c('age', 'lonesc', 'exerc', 'internet_1', 'fincur')

# Unified Data Cleaning (NAs are preserved globally)
df <- hms_data %>%
  mutate(across(c(gender_trans, gender_queer, gender_nonbin, gender_selfID,
                  sexual_h, sexual_l, sexual_g, sexual_bi, sexual_queer,
                  sexual_quest, sexual_pan, sexual_asexual),
                  ~replace_na(., 0))) %>%
  mutate(
      trans = (gender_trans == 1 | gender_queer == 1 | gender_nonbin == 1 | 
               gender_selfID == 1),
      queer_flag = (sexual_l == 1 | sexual_g == 1 | sexual_bi == 1 | 
                    sexual_queer == 1 | sexual_quest == 1 | sexual_pan == 1 | 
                    sexual_asexual == 1),
      identity_group = case_when(
          trans ~ "TGNC",
          !trans & queer_flag ~ "Cis-LGBQ+",
          !trans & !queer_flag & sexual_h == 1 ~ "Cis-Het",
          TRUE ~ NA_character_
      )
  ) %>%
  mutate(
      identity_group = as.factor(identity_group),
      ther_any = as.factor(ther_any),
      queer = as.factor(queer_flag),
      sleep_wknight = as.numeric(sleep_wknight),
      deprawsc = as.numeric(deprawsc),
      flourish = as.numeric(flourish)
  )
```



# Descriptive Plots and Transformations
Our initial correlation plot suggests strong linear relationships between our composite variables. There appears to be a distinct cluster of positive correlations between depression (deprawsc), anxiety (anx_score), and loneliness (lonesc), which are all strongly negatively correlated with flourishing (flourish).

```{r}
pca_data <- df %>% 
  dplyr::select(flourish, deprawsc, anx_score, lonesc, ed_sde) %>%
  drop_na()

corr_matrix <- cor(pca_data)
corrplot.mixed(corr_matrix, lower = "number", upper = "ellipse")
```

We sample $5,000$ observations for our matrix plot, as plotting all $68,062$ observations would result in overplotting and be computationally intensive. We see strong structure and skewness in the Q-Q plot, suggesting that the data are not multivariate normal. At the cost of interpretability, we transform the variables, not only to hopefully attain multivariate normality (which is not required for PCA), but to improve linearity and reduce outlier influence:  

- Square Root: deprawsc and anx_score to correct for right-skew.
- Power Transformation ($x^2$): flourish to correct for left-skew.

```{r}
df_transform <- df %>%
    mutate(
        deprawsc_sqrt = sqrt(deprawsc + 0.5),
        anx_sqrt = sqrt(anx_score + 0.5),
        flourish_sq = flourish^2
    )
```



# Method 1: Principal Component Analysis
Based on these correlations, combined with the aforementioned ratio of $n=68,062$ observations against $p=5$ variables (a response-to-variable ratio of over $13,600:1$), PCA is expected to perform well.  

```{r}
pca_data_transform <- df_transform %>% 
  dplyr::select(flourish_sq, deprawsc_sqrt, anx_sqrt, lonesc, ed_sde) %>%
  drop_na()

pca <- prcomp(pca_data_transform, scale. = TRUE)
screeplot(pca, type = "lines", col = "red", lwd = 2, pch = 19, cex = 1.2, 
          main = "Scree Plot of Transformed Data")
```

Note that the first principal component (PC1) captures $56\%$ of the variance. Adding PC2 increments this to $74\%$ of the variance, and PC3 brings us to $86\%$. However, note that only one eigenvalue is greater than one: $\lambda_{\text{PC1}} = 2.82$. There is a clear elbow after the first principal component, which suggests keeping only PC1, just as suggested by the eigenvalue $>1$ rule.

```{r}
round(pca$rotation[, 1:2], 2)
```

Based on the loadings for PC1, we see two predominant axes emerge: a measure of general well-being and a measure of psychological distress. Observe that flourishing is the only variable with a positive loading ($0.42$), indicating that a higher score on the flourishing scale contributes directly to a higher PC1 score. Conversely, the remaining variables have negative loadings, with depression having the largest effect (-0.53) on a reduced PC1 score. Effectively, we can interpret PC1 to be a measure of a student's psychological health.  



# Method 2: Ordination
While PCA gave us a broad, linear index of psychological health using composite scores, we next use Ordination to map the specific, topological relationships between individual symptoms. We begin with a Correspondence Analysis (CA).

```{r}
set.seed(4747)
df_subset <- df %>% 
  dplyr::select(all_of(symptoms), all_of(environment)) %>%
  drop_na() %>%
  sample_n(500)

hms_symptoms_subset <- df_subset %>% dplyr::select(all_of(symptoms))
hms_env_subset <- df_subset %>% dplyr::select(all_of(environment))

hms_ca <- cca(hms_symptoms_subset)

plot(hms_ca, choices = c(1, 2), type = "n", 
     main = "CA of Symptoms (Dimensions 1 & 2)")
points(hms_ca, display = "sites", col = "lightgrey", pch = 16, cex = 0.5)
text(hms_ca, display = "species", col = "blue", cex = 1, font = 2)
```

We see that our total inertia is 0.1073. Upon visualizing the first two dimensions, there is no clear evidence of an arch effect (data snaking). The lack of snaking and the tight clustering near the origin indicate that mental health symptoms measured by PHQ-9 and GAD-7 are highly comorbid; students experience these symptoms simultaneously rather than turning over from one distinct symptom to another.

Because our symptom scales rely on ordinal frequency with high rates of joint absences, we pivot to Non-Metric Multidimensional Scaling (NMDS) using a Bray-Curtis distance metric to better preserve the rank-order topology.

```{r}
hms_nmds2 <- metaMDS(hms_symptoms_subset, distance = "bray", k = 2, trymax = 50)
```

```{r}
plot(hms_nmds2, type = "n", main = "2D NMDS Solution of HMS Symptoms")
points(hms_nmds2, display = "sites", col = "lightgrey", pch = 16, cex = 0.5)
text(hms_nmds2, display = "species", col = "blue", cex = 1, font = 2)
```

The 2-dimensional plot (Stress = 0.1347, well below the 0.20 threshold) reveals a clear mapping of the symptom space. Observe the extreme observation phq9_9 on the far left-hand side, which specifically measures suicidal ideation and self-harm. A less drastic, but still notable, distinction is the vertical axis, which separates physical symptoms like phq9_3 (sleep) and phq9_4 (fatigue) at a lower elevation from GAD-7 anxiety variables higher up.

We can overlay our environmental continuous variables onto this 2D NMDS solution to see what drives this symptom topology.

```{r}
fit_env <- envfit(hms_nmds2, hms_env_subset, permutations = 999)

plot(hms_nmds2, type = "n", main = "NMDS with Environmental Vectors")
points(hms_nmds2, display = "sites", col = "lightgrey", pch = 16, cex = 0.5)
text(hms_nmds2, display = "species", col = "blue", cex = 0.9, font = 2)
plot(fit_env, col = "red", lwd = 2)
```

Looking at the direction of each arrow, lonesc (loneliness, $R^2 = 0.3228$) and internet_1 point down and to the right, which is where a blend of general anxiety and depression symptoms lie. Meanwhile, fincur (financial stress) pulls hard to the left, acting along the primary axis where the severe outlier phq9_9 sits, indicating it is a major driver behind more severe ideation.



# Method 3: Two-Way MANOVA
With the topology of student mental health and environmental drivers mapped, we investigate how demographic and lifestyle factors independently and interactively influence these outcomes. We run a two-way Type III MANOVA to examine the effect of therapy and sexual orientation on depression and flourishing.

```{r}
df_manova <- df %>% 
  dplyr::select(deprawsc, flourish, ther_any, queer, sleep_wknight) %>%
  drop_na()

manova_fit <- lm(cbind(deprawsc, flourish) ~ ther_any * queer, data = df_manova)
Manova(manova_fit, type = 'III')
```

Using Pillai's Trace, all efforts were statistically significant (p < 0.001). The F statistics reveal that the main effects of sexual orientation (F = 513) and therapy (F = 250) are substantially larger than their interaction (F = 10).

We add average hours of sleep per weeknight (sleep_wknight) as a continuous predictor.

```{r}
model_cov <- lm(cbind(deprawsc, flourish) ~
                ther_any + queer + sleep_wknight, data = df_manova)

Manova(model_cov, type = "III")
```

Somewhat surprisingly, sleep is not only a significant predictor in our multivariate analysis: it is also the largest (F = 1,348). We see that for every additional hour of sleep a student gets, their depression score drops by 1.19 points, and their flourish score increases by 0.98 points.



# Method 4: Discriminant Analysis
Given the significant multivariate differences between demographic groups established in MANOVA, we applied Discriminant Analysis to see if mental health scores alone can predict identity group membership (Cis-Heterosexual, Cis-LGBQ+, and TGNC).

```{r}
df_lda <- df %>% 
  dplyr::select(identity_group, deprawsc, anx_score, lonesc, flourish) %>%
  drop_na()

# Run Cross-Validated LDA
lda_cv <- lda(
    identity_group ~ deprawsc + anx_score + lonesc + flourish,
    data = df_lda,
    prior = c(1/3, 1/3, 1/3),
    CV = TRUE)

lda_cv_acc <- round(sum(diag(prop.table(table(df_lda$identity_group, lda_cv$class)))), 4)
cat("LDA CV Accuracy:", lda_cv_acc, "\n")

# Confusion Matrix
conf_matrix <- table(Actual = df_lda$identity_group, Predicted = lda_cv$class)
conf_matrix
```

Our overall cross-validated accuracy is approximately 52%. The model is most effective at predicting Cis-Het and TGNC, with recalls of 63.3% and 54.4%, accordingly. However, the Cis-LGBQ+ group has a low recall of 19.0%. Most of these students are misclassified as TGNC, illustrating that their psychological health data doesn't form a unique, linearly separable cluster.



# Conclusion and Discussion
Through a comprehensive multivariate approach, we established that student mental health is largely driven by a single latent construct of general distress and well-being (PCA). However, digging deeper into the topology of individual symptoms via Ordination (NMDS) revealed a nuanced structure: while symptoms are highly comorbid, severe clinical manifestations like self-harm map differently than general anxiety, and these are distinctly driven by environmental stressors like financial burden and loneliness. Our MANOVA further contextualized this by revealing that sexual minority status and therapy utilization are strong main effects driving distress, though healthy lifestyle factors like sleep can meaningfully offset it. Finally, Discriminant Analysis confirmed that while mental health profiles differ significantly by identity, the lived experiences of Cis-LGBQ+ and TGNC students share enough psychological overlap that they are difficult to linearly separate, underscoring the shared systemic stressors experienced by the broader LGBTQIA+ umbrella.