---
title: "Homework Set 1"
subtitle: "S&DS 5360 | Multivariate Statistics"
author: "Brandon Tran (bat53)"
date: today
date-format: "MMMM DD, YYYY"
format: 
    pdf:
        geometry: 
            - margin=1in
        colorlinks: true
---



# Problem 1

We have the vectors:

$$
\mathbf{a} = \begin{bmatrix}
    5 & 1 & 14 & 3 & -5 & 12 & -20
\end{bmatrix}
$$

and

$$
\mathbf{b} = \begin{bmatrix}
    0 & 2 & -4 & 10 & 4 & -3 & -4
\end{bmatrix}
$$

Which we store in `R` as:
```{r}
a <- c(5, 1, 14, 3, -5, 12, -20)
b <- c(0, 2, -4, 10, 4, -3, -4)
```


## (a) $\mathbf{a+b}$
```{r}
a+b
```


## (b) $\mathbf{3a-4b}$
```{r}
3*a - 4*b
```


## (c) Length of $\mathbf{a}$
In this case, length denotes the Euclidean Norm, or magnitude $||\mathbf{a}|| = \sqrt{\sum a_i^2}$.
```{r}
sqrt(sum(a^2))
```


## (d) Scalar product $\mathbf{ab}$
```{r}
a %*% b
```


## (e) Interpretation of (d)
Since the scalar (dot) products of $\mathbf{a}$ and $\mathbf{b}$ is $0$, we can conclude that the vectors are orthogonal.



# Problem 2

Consider the vector:
$$
\mathbf{a} = \begin{bmatrix}
    10 & 6 & -5
\end{bmatrix}
$$

Which we store in `R` as:
```{r}
a <- c(10, 6, -5)
```


## (a)
Define the three-dimensional standard basis vectors to be $\mathbf{e}_1 = (1,0,0)$, $\mathbf{e}_2 = (0,1,0)$, and $\mathbf{e}_3 = (0,0,1)$. It follows, then, that 

$$
\mathbf{a} = (10 \times \mathbf{e}_1) + (6 \times \mathbf{e}_2) - (5 \times \mathbf{e}_3)
$$


## (b)
Recall the geometric definition of the dot product:

$$
\begin{aligned}
\mathbf{a} \cdot \mathbf{b} &= ||\mathbf{a}||||\mathbf{b}||\cos(\theta) \\
\cos(\theta) &= \frac{\mathbf{a} \cdot \mathbf{e}_2}{||\mathbf{a}||||\mathbf{e}_2||} \\
&= \frac{6}{||\mathbf{a}||}
\end{aligned}
$$

We use `R` to calcuate this to be:
```{r}
6/sqrt(sum(a^2))
```


## (c)
The subspace spanned by $\mathbf{e}_2$ and $\mathbf{e}_3$ is the yz-plane. The projection of $\mathbf{a}$ onto this plane is therefore $\mathbf{a}_{\text{proj}} = (0, 6, -5)$.



# Problem 3


## (a)
The resulting matrix is a $1 \times 3$ matrix.


## (b)
The resulting matrix is a $12 \times 9$ matrix.


## (c)
These two matrices cannot be multipled, as the vector $\mathbf{a}$ only has one column, which does not match the three rows of $\mathbf{B}$.


## (d)
The resulting matrix is a $1 \times 12$ matrix.



# Problem 4
Recall that a matrix is invertible if its inverse exists. That is, $\exists \mathbf{A}^{-1}$ s.t. $\mathbf{A}\mathbf{A}^{-1} = \mathbf{I}$, which guarantees that the rows and coumns of $\mathbf{A}$ are all linearly independent. For the provided matrix, note that column 3 is the sum of columns 1 and 2. Thus, the matrix is **not invertible.**



# Problem 5
We wish to calculate the characteristic equation and eigenvalues for the following matrix:

$$
\begin{bmatrix}
1 & 3 \\
2 & 4 \\
\end{bmatrix}
$$

Given a constant $\lambda$ and identity matrix $\mathbf{I}$, define

$$
\mathbf{A} - \lambda \mathbf{I} = 
\begin{bmatrix}
1 & 3 \\
2 & 4 \\
\end{bmatrix} -
\lambda \begin{bmatrix}
1 & 0 \\
0 & 1 \\
\end{bmatrix} = 
\begin{bmatrix}
1-\lambda & 3 \\
2 & 4-\lambda \\
\end{bmatrix}
$$

The determinant of this matrix is:

$$
|\mathbf{A} - \lambda \mathbf{I}| = (1-\lambda)(4-\lambda) - 6
$$

Setting this equation to zero and solving for $\lambda$ yields the characteristic equation:

$$
\boxed{\lambda^2-5\lambda-2 = 0}
$$

Solving this characteristic equation yields our two eigenvalues:

$$
\boxed{\lambda = \frac{5}{2} \pm \frac{\sqrt{33}}{2} \approx -0.3723 \text{ or } 5.372}
$$



# Problem 6
Consider points $(1,1), (2,2), (5,5), (-1,-1)$. We want to find a rotation of the space that requires only one dimension to describe the points. Note that these points appear to reside along the line $y=x$ with a $y$-intercept of $0$:

```{r}
df <- matrix(c(1, 1, 2, 2, 5, 5, -1, -1), ncol = 2, byrow = TRUE)
plot(df)
```

By this observation, if we **rotate the points by 45 degrees**, the points will all lie in one dimension. Particularly, if we rotate counter-clockwise by $-45^{\circ}$, the points will lie along the $x$-axis.

Define our rotation matrix to be as follows for $\theta = -45^{\circ}$:

$$
R = 
\begin{bmatrix}
\cos(\theta) & -\sin(\theta) \\
\sin(\theta) & \cos(\theta)
\end{bmatrix} =
\begin{bmatrix}
\cos(-45^{\circ}) & -\sin(-45^{\circ}) \\
\sin(-45^{\circ}) & \cos(-45^{\circ})
\end{bmatrix}
$$

We rotate the points by multiplying our dataframe by the transpose of the rotation matrix defined above:
```{r}
theta <- -45 * (pi / 180)  # Convert to radians.

rot_mat <- matrix(c(cos(theta), -sin(theta),
                    sin(theta), cos(theta)), 
                  nrow = 2, byrow = TRUE)

rotated <- round(df %*% t(rot_mat), 4); rotated
```

Our resulting rotated point matrix is in one dimension, as constructed:

```{r}
plot(rotated)
```



# Problem 7
We are provided two matrices, which we encode into `R`:
```{r}
A <- matrix(c(3,9,12,4,6,0,2,-3,17,-5,3,1,19,19,19,19), nrow = 4, byrow = TRUE)
B <- matrix(c(132,130,156,210,255,255,255,0,130,132,156,1,130,130,1,9),
                                                        nrow = 4, byrow = TRUE)
```


## (a) Transpose of $B$.
```{r}
t(B)
```


## (b) Product of $AB$.
```{r}
A %*% B
```


## (c) Inverse of $A$.
```{r}
solve(A)
```


## (d) Determinant of $A$.
```{r}
det(A)
```


## (e) $AA^{-1}$.
```{r}
A %*% solve(A)
```


## (f) Trace of $B$.
```{r}
sum(diag(B))
```


## (g) Trace of $B^{\text{T}}$.
```{r}
sum(diag(t(B)))
```