---
title: "Homework 5: Cluster Analysis"
subtitle: "S&DS 5360 | Multivariate Statistics"
author: "Brandon Tran (brandon.tran@yale.edu)"
date: today
date-format: "MMMM DD, YYYY"
format: pdf 
---



```{r, include = FALSE}
library(MASS)
library(dplyr)
library(ggplot2)
library(NbClust)
library(haven)
library(tidyr)
library(knitr)
library(kableExtra)
source("https://raw.githubusercontent.com/jreuning/sds363_code/refs/heads/main/HClusEval3.R.txt")
```



# Setup & Data Overview
We continue our work with the Healthy Minds Survey (HMS) data. The variables
are continuous, but lie on different scales due to the construction of each
variable (a sum of underlying Likert-scale questions). Our total number of
observations is $n = 68{,}062$.

- Flourishing (`flourish`): 8–56
- Depression (`deprawsc`): 0–27
- Anxiety (`anx_score`): 0–21
- Loneliness (`lonesc`): 3–9
- Disordered Eating (`ed_sde`): 0–5

We omit the code to preprocess the data.

```{r, echo = FALSE, cache = TRUE}
hms_data <- read_sav("../data/hms_data.sav", encoding = 'latin1')

df <- hms_data %>%
    select(flourish, deprawsc, anx_score, lonesc, ed_sde) %>%
    na.omit()
```



# Part 1 | Distance Metrics, Standardization, and Transformation
Because we are clustering strictly on continuous, numeric mental health summary
scores, standard geometric distance metrics are the most appropriate for our
data. As such, we will elect to use the traditional Euclidean distance
(straight-line distance between students' standardized score profiles). As a
secondary metric, we elect to use Manhattan distance, which is less sensitive
to extreme outliers.

First, recalling our previous analyses on this dataset, we may need to
transform our variables to normalize our data.

```{r}
df %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Score") %>%
  ggplot(aes(x = Score)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  facet_wrap(~ Variable, scales = "free") +
  theme_minimal() +
  labs(title = "Histograms of Mental Health Variables",
       x = "Score", y = "Frequency")
```

The histograms for anxiety, depression, and disordered eating all exhibit right
skews. Flourish exhibits a left skew, and loneliness retains the underlying
ordinal structure from its smaller subset of questions. To correct these, we
propose a log transformation to the right-skewed variables and a square
transformation to the left-skewed variable. We present the resulting Q-Q plots
following the transformations:

```{r, cache = TRUE}
df_transformed <- df %>%
    mutate(
        deprawsc_log = log(deprawsc + 1),
        anx_score_log = log(anx_score + 1),
        ed_sde_log = log(ed_sde + 1),
        flourish_sq = flourish^2) %>%
    select(-deprawsc, -anx_score, -ed_sde, -flourish)

par(mfrow = c(2,3))

df_transformed %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Score") %>%
  ggplot(aes(sample = Score)) +
  stat_qq(color = "black", alpha = 0.5) +
  stat_qq_line(color = "red") + 
  facet_wrap(~ Variable, scales = "free") +
  theme_minimal() +
  labs(title = "Q-Q Plots of Transformed Mental Health Variables",
       x = "Theoretical Normal Quantiles",
       y = "Actual Sample Quantiles")
```

We see a staircase effect in the Q-Q plots, which is inherent to the bounded
survey data we are working with; transformations cannot change this. However,
a bulk of our observations follow the Q-Q line, and observe that our post-log
transformed variables exhibit lighter tails, meaning our outliers were pulled
inward. This ensures that extreme outliers will not disproportionally inflate
the Euclidean distance calculations between students. We see a similar effect
with our squared `flourish` variable.

As can be observed in the overview in the previous section, each continuous
variable exists along a different scale. We will need to standardize our
variables; else, flourishing and other variables with lage ranges will
dominate. To standardize, we convert our scores to Z-scores ($Z \sim N(0,1)$).

```{r}
df_scaled <- scale(df_transformed)
```



# Part 2 | Hierarchical Cluster Analysis
With our data transformed and scaled, we now perform hierarchical cluster
analysis. Given we are working with continuous data, we will compare both
Euclidean and Manhattan distances across Ward's, complete, and average
linkage:

- **Euclidean + Ward's:** The literature supports this as the most balanced
approach for our continuous data. We will use this as our baseline for
comparison with the other methods.
- **Euclidean + Complete Linkage:** Controlling for our distance metric, this
comparison allows us to isolate how changing the agglomeration method (from
minmized variance to furthest neighbor) changes our clusters.
- **Manhattan + Complete Linkage:** Controlling for agglomeration method, we
can see how changing our distance metric impacts the tree.
- **Manhattan + Average Linkage:** Another comparison to further analyze the
outlier-resistant nature of Manhattan distance.

Note that we will need to sample our data to $n = 5,000$ for computational
feasibility. Our true sample size exceeds the maximum limitation for the
`hclust()` function.

```{r, cache = TRUE, fig.width = 8, fig.height = 6, out.width = "100%"}
# Sample n = 5,000
set.seed(4747)
df_sample <- as.data.frame(df_scaled) %>% sample_n(5000)

# Calculate Distance Metrics
dist_eucl <- dist(df_sample, method = "euclidean")
dist_manh <- dist(df_sample, method = "manhattan")

# Run Cluster Analysis
hc_eucl_ward <- hclust(dist_eucl, method = "ward.D")
hc_eucl_comp <- hclust(dist_eucl, method = "complete")
hc_manh_comp <- hclust(dist_manh, method = "complete")
hc_manh_avg <- hclust(dist_manh, method = "average")

# Plot
par(mfrow = c(2, 2))
plot(hc_eucl_ward, labels = FALSE, cex = 0.5, 
     xlab = "", ylab = "Distance", main = "Euclidean + Ward's", sub = "")
plot(hc_eucl_comp, labels = FALSE, cex = 0.5, 
     xlab = "", ylab = "Distance", main = "Euclidean + Complete", sub = "")
plot(hc_manh_comp, labels = FALSE, cex = 0.5, 
     xlab = "", ylab = "Distance", main = "Manhattan + Complete", sub = "")
plot(hc_manh_avg, labels = FALSE, cex = 0.5, 
     xlab = "", ylab = "Distance", main = "Manhattan + Average", sub = "")
```

The resulting dendrograms visibly demonstrate how different agglomeration
methods handle dense survey data like ours. Euclidean + Ward suggests between
3–5 distinct clusters. When we switch to complete linkage, instead of forming
distinct tree trunks, the data clumps into a massive, dense blob, with small
individual branches very close to the top.  Generally, complete and average
linkage methods were difficult to interpret, suffering from severe chaining and
highly unbalanced tree structures with poorly defined separation at higher
distances. Meanwhile, Euclidean distance paired with Ward's method produced a
highly intepretable denderogram, minimizing within-cluster variance and
separating the data into distinct, balanced branches.



# Part 3 | Discussion on Groups
To confirm the 3–5 clusters that we visually assessed from the Euclidean + Ward
analysis in Part 2, we mathematically evaluate the number of clusters needed
via `NbClust` and the courses's custom `HClusEval`. We perform this on a small
sample of data ($n = 1{,}000$) for computational purposes.

```{r, cache = TRUE, message = FALSE, warning = FALSE}
set.seed(4747)
df_sample <- as.data.frame(df_scaled) %>% sample_n(1000)

par(mfrow = c(1, 2))

eval <- hclus_eval(df_sample, dist_m = 'euclidean', clus_m = 'ward',
           plot_op = T, print_num = 10)

nb_results <- NbClust(df_sample, distance = 'euclidean', min.nc = 2,
                      max.nc = 10, method = 'ward.D', index = 'all',
                      alphaBeale = 0.1)
```

Visual inspection of the evaluation plots reveal a distinct elbow in the R-
Squared (RSQ) trajectory at $k = 4$ clusters, with diminishing returns in 
explained variance beyond this point. The Hubert statistic and D-index second
difference plots also exhibited significant peaks at $k = 4$. The `NbClust`
majority rule confirms our visual assessment, concluding that $k = 4$ is the
optimal number of clusters based on $11$ voting indices. For reference, a
visualization of these four clusters is plotted below:


```{r}
par(mfrow = c(1, 1))
cuts <- cutree(hc_eucl_ward, k = 4)
plot(hc_eucl_ward, labels = FALSE, cex = 0.6, xlab = "", ylab = "Distance",
     main = "Clustering of HMS Students", sub = '')
rect.hclust(hc_eucl_ward, k = 4)
```

We proceed with $k = 4$ for our final $k$-means partitioning.


# Part 4 | $k$-Means
We perform $k$-means clustering on our data and compare with our results from
Part 3. Note that to ensure model convergence with our large sample size, we
use the MacQueen algorithm with $50$ iterations.

```{r, warning = FALSE, message = FALSE}
set.seed(4747)

wss <- numeric(10) 

for (k in 1:10) {
  wss[k] <- kmeans(df_scaled, centers = k, nstart = 25,
                   iter.max = 50, algorithm = "MacQueen")$tot.withinss
}

plot(1:10, wss, type = "b", pch = 19, col = "blue", lwd = 2,
     xlab = "Number of Clusters (k)", 
     ylab = "Total Within-Cluster Sum of Squares (WSS)",
     main = "K-Means Elbow Method")

final_kmeans <- kmeans(df_scaled, centers = 4, nstart = 25,
                       iter.max = 50, algorithm = "MacQueen")

df_final <- df %>%
  mutate(Cluster = as.factor(final_kmeans$cluster))
```

Our Total Within-Cluster Sum of Squares (WSS) plot exhibits a distinct elbow
at $k = 4$, thereafter which the steep reduction in variance begins to plateau.
When comparing these results to the hierarchical analysis in Step 3, the
findings are consistent (though an elbow at $k=2$ could also be discerned in
this method). Thus, both the objective mathematical indices and the visual $k$-
means elbow plot independently suggest $k = 4$ natural groupings within our
data.



# Steps 5–6 | Visualization in PCA Space
Because the data we are working with exists in $5$-dimensional space, we
collapse our data in to just two dimensions to help us visualize our groupings.
While we could use both PCA and discriminant analysis, both methods of which
we've previously used on this dataset, we elect to use PCA because it is an
unsupervised method, whereas discirminant analysis is supervised. That is, PCA
will serve as an unbiased canvas for us to visualize our groups on.

```{r}
pca <- prcomp(df_scaled)

pca_data <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  Cluster = as.factor(final_kmeans$cluster)
)

ggplot(pca_data, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(alpha = 0.1, size = 1) + 
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 4))) + 
  theme_minimal() +
  labs(title = "K-Means Clusters Visualized via PCA (k=4)",
       x = "Principal Component 1",
       y = "Principal Component 2",
       color = "Cluster")
```

We see four distinct clusters in along our two-dimensional principal axes,
confirming our previous analyses. While some boundary overlap exists, this is
expected; the clusters themselves are largely continuous with dense central
hotspots. We do observe some striping within the clusters, notably cluster #3
(in blue), but this is likely an artifact of the discrete, bounded nature of
the underlying Likert-scaled survey data.



# Step 7 | Final Analysis of Groups
With multiple independent analyses confirming that $k=4$ groups exist within
our data, we endeavor to assign a meaning to each group. To do this, we present
the average scores per variable, per cluster:

```{r}
cluster_profiles <- df_final %>%
  group_by(Cluster) %>%
  summarise(
    Count = n(), 
    Mean_Flourish = round(mean(flourish, na.rm = TRUE), 2),
    Mean_Depress  = round(mean(deprawsc, na.rm = TRUE), 2),
    Mean_Anxiety  = round(mean(anx_score, na.rm = TRUE), 2),
    Mean_Eating   = round(mean(ed_sde, na.rm = TRUE), 2),
    Mean_Lonely   = round(mean(lonesc, na.rm = TRUE), 2)
  )
```

```{r, echo = FALSE}
kable(cluster_profiles, 
      format = "latex",
      booktabs = TRUE,
      col.names = c("Cluster", "Student Count", "Flourishing", 
                    "Depression", "Anxiety", "Eating Concerns", "Loneliness"),
      align = "c",
      caption = "Final K-Means Cluster Profiles (k=4)") %>%
  kable_styling(latex_options = c("striped", "HOLD_position"), 
                full_width = FALSE,
                font_size = 10) %>%
  row_spec(0, bold = TRUE)
```

Most notably, Cluster 4 stands out as our thriving baseline, exhibiting the
highest flourishing score and, conversely, extremely low scores across all
negative psychological variables. Meanwhile, Cluster 2 seems to represent our
most high-risk population, with the lowest flourishing score and highest
scores across negative psychological variables. Cluster 3 is unique: while at
first glance, these students appear to be high-functioning, exhibiting high
flourishing scores and relatively low depression and anxiety, their eating
concern score is dramatically elevated compared to baseline. Thus, they likely
form an at-risk group unique to disordered eating. Finally, Cluster 1 seems to
be comprised of the opposite: relatively elevated depression, anxiety, and
loneliness, but with the lowest eating concern score. This seems to indicate
a population dealing with generalized, sub-clinical internalizing symptoms who
do not engage in disordered eating behaviors.