---
title: "Homework 5: Cluster Analysis"
subtitle: "S&DS 5360 | Multivariate Statistics"
author: "Brandon Tran (brandon.tran@yale.edu)"
date: today
date-format: "MMMM DD, YYYY"
format: pdf 
---



```{r, include = FALSE}
library(MASS)
library(dplyr)
library(tidyr)
library(haven)
library(car)
library(emmeans)
library(ggplot2)
library(patchwork)
library(heplots)
library(vegan)
```



# Setup & Data Overview
We continue our work with the Healthy Minds Survey (HMS) data. The variables
are continuous, but lie on different scales due to the construction of each
variable (a sum of underlying Likert-scale questions). Our total number of
observations is $n = 68{,}062$.

- Flourishing (`flourish`): 8–56
- Depression (`deprawsc`): 0–27
- Anxiety (`anx_score`): 0–21
- Loneliness (`lonesc`): 3–9
- Disordered Eating (`ed_sde`): 0–5

We omit the code to preprocess the data.

```{r, echo = FALSE, cache = TRUE}
hms_data <- read_sav("../data/hms_data.sav", encoding = 'latin1')

df <- hms_data %>%
    select(flourish, deprawsc, anx_score, lonesc, ed_sde) %>%
    na.omit()
```



# Part 1 | Distance Metrics, Standardization, and Transformation
Because we are clustering strictly on continuous, numeric mental health summary
scores, standard geometric distance metrics are the most appropriate for our
data. As such, we will elect to use the traditional Euclidean distance
(straight-line distance between students' standardized score profiles). As a
secondary metric, we elect to use Manhattan distance, which is less sensitive
to extreme outliers.

First, recalling our previous analyses on this dataset, we may need to
transform our variables to normalize our data.

```{r}
df %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Score") %>%
  ggplot(aes(x = Score)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  facet_wrap(~ Variable, scales = "free") +
  theme_minimal() +
  labs(title = "Histograms of Mental Health Variables",
       x = "Score", y = "Frequency")
```

The histograms for anxiety, depression, and disordered eating all exhibit right
skews. Flourish exhibits a left skew, and loneliness retains the underlying
ordinal structure from its smaller subset of questions. To correct these, we
propose a log transformation to the right-skewed variables and a square
transformation to the left-skewed variable. We present the resulting Q-Q plots
following the transformations:

```{r, cache = TRUE}
df_transformed <- df %>%
    mutate(
        deprawsc_log = log(deprawsc + 1),
        anx_score_log = log(anx_score + 1),
        ed_sde_log = log(ed_sde + 1),
        flourish_sq = flourish^2) %>%
    select(-deprawsc, -anx_score, -ed_sde, -flourish)

par(mfrow = c(2,3))

df_transformed %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Score") %>%
  ggplot(aes(sample = Score)) +
  stat_qq(color = "black", alpha = 0.5) +
  stat_qq_line(color = "red") + 
  facet_wrap(~ Variable, scales = "free") +
  theme_minimal() +
  labs(title = "Q-Q Plots of Transformed Mental Health Variables",
       x = "Theoretical Normal Quantiles",
       y = "Actual Sample Quantiles")
```

We see a staircase effect in the Q-Q plots, which is inherent to the bounded
survey data we are working with; transformations cannot change this. However,
a bulk of our observations follow the Q-Q line, and observe that our post-log
transformed variables exhibit lighter tails, meaning our outliers were pulled
inward. This ensures that extreme outliers will not disproportionally inflate
the Euclidean distance calculations between students. We see a similar effect
with our squared `flourish` variable.

As can be observed in the overview in the previous section, each continuous
variable exists along a different scale. We will need to standardize our
variables; else, flourishing and other variables with lage ranges will
dominate. To standardize, we convert our scores to Z-scores ($Z \sim N(0,1)$).

```{r}
df_scaled <- scale(df_transformed)
```



# Part 2 | Hierarchical Cluster Analysis
With our data transformed and scaled, we now perform hierarchical cluster
analysis. Given we are working with continuous data, we will compare both
Euclidean and Manhattan distances across Ward's, complete, and average
linkage:

- **Euclidean + Ward's:** The literature supports this as the most balanced
approach for our continuous data. We will use this as our baseline for
comparison with the other methods.
- **Euclidean + Complete Linkage:** Controlling for our distance metric, this
comparison allows us to isolate how changing the agglomeration method (from
minmized variance to furthest neighbor) changes our clusters.
- **Manhattan + Complete Linkage:** Controlling for agglomeration method, we
can see how changing our distance metric impacts the tree.
- **Manhattan + Average Linkage:** Another comparison to further analyze the
outlier-resistant nature of Manhattan distance.

Note that we will need to sample our data to $n = 5,000$ for computational
feasibility. Our true sample size exceeds the maximum limitation for the
`hclust()` function.

```{r}
# Sample n = 5,000
set.seed(4747)
df_sample <- as.data.frame(df_scaled) %>% sample_n(5000)

# Calculate Distance Metrics
dist_eucl <- dist(df_sample, method = "euclidean")
dist_manh <- dist(df_sample, method = "manhattan")

# Run Cluster Analysis
hc_eucl_ward <- hclust(dist_eucl, method = "ward.D")
hc_eucl_comp <- hclust(dist_eucl, method = "complete")
hc_manh_comp <- hclust(dist_manh, method = "complete")
hc_manh_avg <- hclust(dist_manh, method = "average")

# Plot
par(mfrow = c(2, 2))
plot(hc_eucl_ward, labels = FALSE, cex = 0.5, 
     xlab = "", ylab = "Distance", main = "Euclidean + Ward's", sub = "")
plot(hc_eucl_comp, labels = FALSE, cex = 0.5, 
     xlab = "", ylab = "Distance", main = "Euclidean + Complete", sub = "")
plot(hc_manh_comp, labels = FALSE, cex = 0.5, 
     xlab = "", ylab = "Distance", main = "Manhattan + Complete", sub = "")
plot(hc_manh_avg, labels = FALSE, cex = 0.5, 
     xlab = "", ylab = "Distance", main = "Manhattan + Average", sub = "")
```

The resulting dendrograms visibly demonstrate how different agglomeration
methods handle dense survey data like ours. Euclidean + Ward suggests between
3–5 distinct clusters. When we switch to complete linkage, instead of forming
distinct tree trunks, the data clumps into a massive, dense blob, with small
individual branches very close to the top.  Generally, complete and average
linkage methods were difficult to interpret, suffering from severe chaining and
highly unbalanced tree structures with poorly defined separation at higher
distances. Meanwhile, Euclidean distance paired with Ward's method produced a
highly intepretable denderogram, minimizing within-cluster variance and
separating the data into distinct, balanced branches.