---
title: "Homework 6: Ordination"
subtitle: "S&DS 5360 | Multivariate Statistics"
author: "Brandon Tran (brandon.tran@yale.edu)"
date: today
date-format: "MMMM DD, YYYY"
format: pdf 
---



```{r, include = FALSE}
library(MASS)
library(dplyr)
library(ggplot2)
library(haven)
library(tidyr)
library(knitr)
library(kableExtra)
library(tidyverse)
library(vegan)
library(vegan3d)
library(mgcv)
```



# Setup & Data Overview
We continue our work with the Healthy Minds Survey (HMS) data. For
Correspondence ANalysis and NMDS, we will select related survey questions
measured on the same scale. For these, we select:

- Depression (`phq9_1` to `phq9_9`): Nine questions on interest, mood, sleep, and energy levels over 2 weeks.
- Anxiety (`gad7_1` to `gad7_7`): Seven questions on nervousiness, worrying, and restlessness over 2 weeks.

For Canonical Correspondence Analysis (CCA) and non-linear wireplots, we select
continuous variables that may explain the variation in symptoms:

- Age (`age`): Continuous measure of respondent's age.
- Loneliness Score (`lonesc`): Sum of UCLA Loneliness Scale (3–9)
- Exercise Hours (`exerc`): Average hours per week spent exercising (1–5 scale).
- Internet Use (`internet_1`): Time spent online (1–8).
- Financial Stress (`fincur`): Stressfulness of current financial situation (1-5 scale).

We omit the code to preprocess the data.

```{r, echo = FALSE, cache = TRUE}
hms_data <- read_sav("../data/hms_data.sav", encoding = 'latin1')

symptoms <- c(paste0('phq9_', 1:9), paste0('gad7_', 1:7))
environment <- c('age', 'lonesc', 'exerc', 'internet_1', 'fincur')

df <- hms_data %>%
    select(all_of(symptoms), all_of(environment)) %>%
    na.omit()

hms_symptoms <- df %>% select(all_of(symptoms))
hms_env <- df %>% select(all_of(environment))
```



# Parts 1–4 | Correspondence Analysis
We begin by performing a an indirect ordination to explore the underlying
structure of mental health symptoms among students, particularly to reduce
the dimensionality of our 16 symptom variables across `PHQ-9` and `GAD-7` to
more easily visualize how they group together. To do so, we complete a
Correspondence Analysis, which is ideal for our data because the symptom
data represents frequency counts (scales from 1 = "Not at All" to 4 = "Nearly
Every Day"). CA maps both students and symptoms into low dimensional space via
chi-square distances.

```{r}
hms_ca <- cca(hms_symptoms)
summary(hms_ca)
```

We see that our total inertia is $0.1073$, representing the total varaince in
our data (calculated as a scaled chi-square statistic). Axis 1 (CA1) explains
$16.48\%$ of the variance, while Axis 2 (CA2) explains $10.56\%$, or together,
$27.04\%$ of the variance in total.

Observe on our two-dimensional plot of Axes 1 and 2 that our clustered symptoms
in blue text tightly ovelap near the origin $(0, 0)$. Because there are so many
students pulling the axes outward, the variation among symptoms themselves make
them appear clumped in the middle. We provide a zoomed-in view on the right-
hand size to more easily discern the symptoms.

```{r, fig.height = 4, fig.width = 8, out.width = "100%"}
par(mfrow=c(1,2))

# Full Plot
plot(hms_ca, choices = c(1, 2), type = "n", 
     main = "CA of Symptoms (Dimensions 1 & 2)")
points(hms_ca, display = "sites", col = "lightgrey", pch = 16, cex = 0.5)
text(hms_ca, display = "species", col = "blue", cex = 1, font = 2)

# Zoomed-In Plot
plot(hms_ca, choices = c(1, 2), type = "n", 
     xlim = c(-.5, .5), ylim = c(-.5, .5),
     main = "Zoomed In")
points(hms_ca, display = "sites", col = "lightgrey", pch = 16, cex = 0.5)
text(hms_ca, display = "species", col = "blue", cex = 1, font = 2)
```

Upon visualizing the first two dimensions of the Correspondence Analysis, we
see there is no clear evidence of an arch effect (data snaking). Rather than
forming a classic U or V crescent shape, the data forms a dense, continuous
cloud centered around the origin, with the symptom variables tightly clustered
in the middle.

As discussed in class, this lack of snaking would be expected when there is
complete species turnover along an environmental gradient. Our mental health
symptoms do not behave this way; rather than "turning over," symptoms are highly
correlated and additive. For example, a student experiencing severe depression
does not replace their more mild symptoms (like fatigue) with more severe ones
(like self harm). Rather, they may experience both simultaneously. Since these
symptoms co-occur, the CA maps them as the tightly inter-corrleated cluster we
see in the plots.

Given the above, we conclude that mental health symptoms measured by `PHQ-9`
and `GAD-7` are highly comorbid and do not represent distinct, mutually
exclusive conditions among the student population. The lack of data snaking and
the tight clustering of symptom variables near the origin indicate that students
do not experience a turnover symptoms. Instead, the first two axes, which
capture $\approx 27 \%$ of the variance, likely represent a single, dominant
gradient of overall psychological distress. Students scale along this gradient
additively, experiencing these anxiety and depression symptoms simultaneously
rather than in isolation of each other.



# Parts 5–7 | Multidimensional Scaling
Now, we perform Non-Metric Multidimensional Scaling (NMDS) in one, two, and
three dimensions. Unlike Correspondence Analysis, which relies on chi-square
distances, NMDS's rank-based approach maps respondents into a low-dimensional
space such that the distances between points optimally preserve the rank-order
of their actual dissimilarites in the full 16-dimensional symptom space.

We elect to use the Bray-Curtis distance metric (as opposed to Euclidean
distance) because of its robustness to joint absences. For example, if two
students both select "not at all" for a number of severe symptoms, Bray-Curtis
does not treat this shared absence as a strong indicator of similarity, focusing
instead on the actually experienced symptoms. This is optimal for survey data
like ours.

```{r, cache = TRUE, message = FALSE, results = FALSE, warning = FALSE}
set.seed(4747)

df_subset <- df %>% sample_n(500)

hms_symptoms_subset <- df_subset %>% select(all_of(symptoms))
hms_env_subset <- df_subset %>% select(all_of(environment))

hms_nmds1 <- metaMDS(hms_symptoms_subset, distance = "bray", k = 1, trymax = 50)
hms_nmds2 <- metaMDS(hms_symptoms_subset, distance = "bray", k = 2, trymax = 50)
hms_nmds3 <- metaMDS(hms_symptoms_subset, distance = "bray", k = 3, trymax = 50)
```

Note that we manually set `trymax = 50` to ensure the algorithm incorporates
enough random starts to reliably find the global optimum within a reasonable
number of runs. Furthermore, we sample $500$ rows from our data, as our datset
containing $33{,}100$ rows of symptom data is comptuationally expensive and
unncessary for ordination.

```{r, fig.height = 4, fig.width = 8, out.width = "100%"}
par(mfrow = c(1,1))

# 1D
plot(hms_nmds1, main = "1D NMDS Solution of HMS Symptoms")

# 2D
plot(hms_nmds2, type = "n", main = "2D NMDS Solution of HMS Symptoms")
points(hms_nmds2, display = "sites", col = "lightgrey", pch = 16, cex = 0.5)
text(hms_nmds2, display = "species", col = "blue", cex = 1, font = 2)

# 3D
pl <- ordiplot3d(hms_nmds3, display = "sites", pch = 16, col = "lightgrey",
                 main = "3D NMDS Solution")
sp <- scores(hms_nmds3, choices = 1:3, display = "species")
text(pl$xyz.convert(sp), rownames(sp), col = "blue", font = 2, cex = 1)
```

In the above plots, we see:
- **1D NMDS:** The 1-dimensional plot forces all symptom variance onto a single axis, resulting in an overcrowded plot from which it is difficult to discern meaningful relationships between individual symptoms.
- **2D NMDS:** The 2-dimensional plot reveals a much clearer mapping of the symptom space. Observe the extreme observation `phq9_9` on the far left-hand side; this question specifically measures suicidal ideation and self-harm. The remaining general anxiety and depression symptms form a looser cluster on the right. The visual separation makes sense: self harm is distinctly different from other, less severe depressive symptoms and requires acute care. A less drastic, but still notable, distinction is the vertical axis, which separates symptoms by clinical nature. Note that `phq_3` (sleep), `phq_4` (fatigue), and `phq_5` (appetite) cluster together at a distinctly lower elevation on the second axis. These questions represent the pysical manifestations of depression. In contrast, the items situated higher on the y-axis are predominantly `GAD-7` anxiety variables, such as uncontrollable worry and feeling on edge. This indicates that the NMDS algorithm successfully captured the latent structural divide between physical exhaustion and depressive symptoms compared to anxious states among the student population.
- **3D NMDS:** The 3-dimensional plot adds an additional axis for depth which is not immmediately helpful. While it helps pull apart hte main cluster of symptoms that overlap in 2D, the overall topology remains the same, with `phq9_9` retaining the same distinct separation.

```{r}
# Print stress numbers
stress_1d <- hms_nmds1$stress
stress_2d <- hms_nmds2$stress
stress_3d <- hms_nmds3$stress

cat("Stress for 1D:", stress_1d, "\n")
cat("Stress for 2D:", stress_2d, "\n")
cat("Stress for 3D:", stress_3d, "\n")

# Plot
dimensions <- c(1, 2, 3)
stress_values <- c(stress_1d, stress_2d, stress_3d)

plot(dimensions, stress_values, type = "b", pch = 19, col = "blue", lwd = 2,
     xlab = "Number of Dimensions (k)", ylab = "Stress",
     main = "NMDS Scree Plot for HMS Symptoms",
     ylim = c(0, max(stress_values) + 0.05), xaxt = "n")

axis(1, at = c(1, 2, 3))

abline(h = 0.20, col = "red", lty = 2)
text(2.5, 0.21, "Acceptable Threshold (0.20)", col = "red", cex = 0.8)
```

The stress of our 1D, 2D, and 3D NMDS solutions were $0.2000$, $0.1347$, and
$0.1057$, respectively. As visualized in the scree plot, moving from a 1D to 2D
solution provides a notable and steep reduction in stress, falling below the
acceptable range of $0.20$. This means that the 2D plot provides a reliable
geometric representation of the symptom distances without severe distortion.
While the 3D solution reduces the stress further, the 2D solution is optimal
for this analysis as it perfectly balances an acceptable goodness-of-fit with
visual interpretability, capturing the primary structural divides in the data.
This confirms what we found in our visual analysis earlier.



# Part 8 | Overlay of Continuous Variables
Proceeding with our 2-dimensional NMDS solution, we overlay the five continous
variables we selected at the start: age, loneliness score, exercise hours, 
internet use, and financial stress.

```{r}
fit_env <- envfit(hms_nmds2, hms_env_subset, permutations = 999)
fit_env
```

Immediately notice that four of our five variables are statistically singificant:
`lonesc`, `exerc`, `internet_1`, and `fincur` (just not `age`). Looking at the
corresponding R-squared values, observe that loneliness has an extremely high
value of $0.3228$, follwed in magnitude by `fincur`, `internet1`, and finally,
`exercise.` This tells us that loneliness, particularly, is a huge driver of
where students land in the symptom space.

```{r}
plot(hms_nmds2, type = "n", main = "NMDS with Environmental Vectors")
points(hms_nmds2, display = "sites", col = "lightgrey", pch = 16, cex = 0.5)
text(hms_nmds2, display = "species", col = "blue", cex = 0.9, font = 2)
plot(fit_env, col = "red", lwd = 2)
```

A visual plot in two dimensions confirms our numerical analysis . We note that
the length of the vectors for each continuous variable match the magnitude of
the aforementioned correlations. Now, observing the direction of each arrow, we
see that `lonesc` and `internet_1` point down and to the right, which is where
a blend of anxiety and depression symptoms lie. Meanwhile, `fincur` pulls hard
to the left, acting along the primary axis where the severe outlier `phq9_9`
sits. This seems to indicate that while loneliness is a major driver behind
anxiety and depression, financial stress, in particular, seems to be a major
driver behind more severe suicidal ideation.

```{r}
plot(hms_nmds2, type = "n", main = "NMDS Wireplot: Loneliness Gradient")
points(hms_nmds2, display = "sites", col = "lightgrey", pch = 16, cex = 0.5)
text(hms_nmds2, display = "species", col = "blue", cex = 0.9, font = 2)
surf_lone <- ordisurf(hms_nmds2, hms_env_subset$lonesc, add = TRUE, col = "darkgreen", lwd = 2)
```



```{r}
vis.gam(surf_lone, main = "3D Surface of Loneliness over NMDS Space", 
        theta = -45, phi = 30, color = "topo")
```